Executive Constraints and Civil Conflict Onset

### 2019-11-03 paper.Rmd
WIP: continue theory section

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" (cite Hodgson). More informally, we can consider
them to be the "rules of the game", encompassing for example codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents.^[None of which is meant to
imply that institutions stand external to actor perceptions and
preferences as unchanging features. Rather, institutions are social
constructs subject to changing normative values and able to be
reinforced or undermined by behavioural feedback loops. This latter
point is especially interesting within the context of civil conflict
considering how the development of various institutions and conflict
incidence can be jointly determined; however, it remains outside the
scope of this paper.] Furthermore, over time, the habits formed from
these systems create expectations of stable patterns of behaviour
thereby shaping the interactions between actors by reducing
uncertainty (cite North). Although there is a wide variety of
institutions, the present focus will be on the type of institutions
which govern political behaviour.

@@

Unsurprisingly, insofar as the role of institutions has been examined
by previous literature the focus has largely been on the broader
notion of democracy, which at a minimum is a system regulating access
to political power through popular representation determined by free
and competitive elections (Schumpter). Democracy has been hypothesised
to have a pacifying domestic effect according to two predominant
mechanisms. The first is centered on the concept of relative
deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence (Gurr 1970). [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means (Hegre).

@@

>>>>

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" (cite Hodgson). More informally, we can consider
them to be the "rules of the game", encompassing for example codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents. Furthermore, over time, the
habits formed from these systems create expectations of stable
patterns of behaviour thereby shaping the interactions between actors
by reducing uncertainty (cite North). Although there is a wide variety
of institutions, the present focus will be on the type of institutions
which govern political behaviour.

[^institutions]: None of which is meant to imply that institutions
    stand external to actor perceptions and preferences as unchanging
    features. Rather, institutions are social constructs subject to
    changing normative values and able to be reinforced or undermined
    by behavioural feedback loops. This latter point is especially
    interesting within the context of civil conflict considering how
    the development of various institutions and conflict incidence can
    by jointly determined; however, it remains outside the scope of
    this paper.

@@

Unsurprisingly, insofar as the role of institutions has been examined
by previous literature the focus has largely been on the broader
notion of democracy, which at a minimum is a system regulating access
to political power through popular representation determined by free
and competitive elections (Schumpter). Democracy has been hypothesised
to have a pacifying domestic effect according to two predominant
mechanisms. The first is centered on the concept of relative
deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence (Gurr 1970). [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means (Hegre). The second mechanism highlights the risk
of future shifts in the distribution of power in undermining the
attempt to reach a bargained solution to avoid war between two parties
in conflict. More specifically, a future change in favor of the
relative strength of one of the parties can provide an incentive to
later renege. This influences the present degree to which concessions
can be offered and parties are willing to credibly commit
(Powell). Within the domestic sphere, democracy offers a possible
solution by formalising the regular contest of power with full
enfranchisement, thereby providing a *de jure* guarantee for the
balance of power (Acemoglu and Robinson), and should, likewise,
decrease the likelihood of civil conflict.

@@

Despite the theoretical basis linking democracy to reduced risks for
civil conflict, the empirical results tend to be mixed. Hegre *et al*
(2001), for example, find that democracies are at a lower risk for
civil conflict; however, no more so than fully authoritarian
states. Instead, anocratic regimes in between democracy and
authoritarianism are considered to be the most prone. The explanatory
logic is that both democratic and fully authoritarian states tend to
be consolidated regimes. In the case of the former, this means the
presence of institutions conducive to conflict resolution, and in the
case of the latter, the capacity to violently repress
dissent. Anocratic regimes are by their very nature transitory, partly
open and partly repressive, encouraging political discontent against
the state without the capacity to fully suppress opposition.

In contrast, studies emphasizing the importance of political and
economic opportunities for rebellion in understanding when and where
conflicts break out, such as Fearon and Laitin (2003) and Collier,
Hoeffler, and Rohner (2009), find democracy to be insignificant. In
similar fashion, democracy is associated with higher levels of
political rights, respect for ethnic and religious minorities, and
less repression, thus being a proxy for the latent level of political
discontent against the state. However, even if the number of political
grievances vary across regime types, they are still ubiquitous enough
to be a poor explanation for why conflict breaks out in only a
relatively small subset of countries. Instead, the argument follows
that the structural conditions conducive to rebellion are the
determining factors.[^opportunities]

[^opportunities]: For Fearon and Laitin these structural conditions
    are related to the features which make rebellion militarily
    viable.  Specifically, there's a higher risk of conflict onset
    where state capacity is weak, principally understood in terms of
    state revenue and its impact on policing and counterinsurgency
    operations; large impoverished populations; and insurgents have a
    terrain advantage. Collier *et al* start from the premise that war
    is expensive and the primary impediment to a rebellion is its
    financial feasibility relative to the strength of the state. On
    the basis of their results they go so far as provide support to
    the proposition, "where a rebellion is financially and militarily
    feasible it will occur." Both approaches, however, reach the same
    conclusion concerning the empirical evidence surrounding the role
    of democracy.

Subsequent research
